---
title: "05_analysis_1.qmd"
editor: visual
format:
  html:
    embed-resources: true
author: "Sarah Rosenberg Asmussen (s194689),	Mette Bøge Pedersen (s194679),	Caroline Amalie Bastholm Jensen (s213427), Jaime Noguera Piera (s233773), Yassine Turki (s231735)"
---

## Library Load

```{r}
library(tidyverse)
library(readr)
```

## Data Load

```{r}
data <- read_tsv("../data/03_dat_aug.tsv")
data
```

## Analysis 1

In the first analysis, we want to identify which genes are found to be significantly different expressed in patients with metastatic cancer compared to non-metastatic cancer. Furthermore, we want to investigate if the gene expression is up-regulated of down-regulated.

The significance was calculated on the basis of a Student's T-test where the expression of each gene was compared based on if the patients had metastasis or not.

The Log2 Fold Change for each gene was calculated based on the average gene expression level by comparing samples with metastasis and no metastasis.

These results are shown in a volcano plot:

```{r}
volcano_plot <- data |> 
  select(gene, log2_fold_change_avg, p_value) |>
  unique() |>
  mutate(log_10_p = -log10(p_value),
         Significance = case_when(p_value > 0.01 ~ "Not significant",
                                    p_value <= 0.01 ~ "Significant")) |> 
  ggplot(mapping = aes(x = log2_fold_change_avg,
                       y = log_10_p,
                       color = Significance)) +
  geom_point(size = 1, alpha = 0.5) +
  geom_hline(yintercept=2,
             linetype="dotted", 
             color = "black", 
             size=0.5) +
  theme(legend.position = "none") + 
  theme_minimal() +
  labs(title="Genes Associated with Metastasis in Bladder Cancer", 
       subtitle = "Genes highlighted in turquoise are significant on a significance level of 0.01",
     x = "Log2 Fold Change",
     y = "-log10(p)") 

ggsave(
  filename = "../results/05_volcano_plot.png",
  plot = volcano_plot,
  device = "png",
  height = 5,
  dpi = 300  
)


print(volcano_plot)
dev.off()
```
